Rapid Epidemiological Analysis of Comorbidities and Treatments as risk factors in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population 

### 2020-05-22 casecontrol.Rmd
fixed to keep memory usage down, also to add obsolete SMR ethnicity codes

---
title: 'Rapid Epidemiological Analysis of Comorbidities and Treatments in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population [**Draft, not to be quoted**]'
author: 
  - name: Paul M McKeigue
    email: paul.mckeigue@ed.ac.uk
    affiliation: Usher
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart J McGurnaghan
    affiliation: IGMM
  - name: Sharon Kennedy
    affiliation: HPS
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Thomas M Caparrotta 
    affiliation: IGMM
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: Helen M Colhoun
    affiliation: IGMM
address: 
   - code: Usher
     address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care
   - code: IGMM
     address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology. 
   - code: Glasgow
     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
   - code: HPS
     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health. 
geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
#    toc: true
#    toc_depth: 2
csl: ../plos.csl
urlcolor: red

@@

---

@@

_Methods_ -- The design was a matched case-control study. Test-positive cases were ascertained as all those with a positive nucleic acid test for SARS-CoV-2 in the national database and from national mortality data.  Those who had been admitted to intensive care or had died within 28 days of a positive test were classified as severe cases.  For each case, ten controls matched for sex, age and primary care practice were selected from the population register.  For cases and controls we extracted all diagnostic codes from the past five years of hospital discharge records, all drug codes from encashed prescriptions issued in primary care care during the past year and linked to the national diabetes register.  Rate ratios for associations with severe COVID-19 were estimated by conditional logistic regression.  The performance of a predictive model for severe COVID-19 was estimated by cross-validation.

_Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]` .  Rate ratios (95% CIs) for conditions previously designated as conferring high risk were  
`r table.agegr.all[4, 3]` for diabetes, 
`r table.agegr.all[8, 3]` for chronic renal failure, 
`r table.agegr.all[5, 3]` for ischemic heart disease,
`r table.agegr.all[6, 3]` for other heart disease, 
`r table.agegr.all[9, 3]` for neurological disease, 
`r table.agegr.all[10, 3]` for chronic liver disease and 
`r table.agegr.all[11, 3]` for immune deficiency or suppression.

Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of these listed conditions.  In those aged under 40 years the frequencies of any listed condition were `r gsub("\\\\", "", table.agegr[13, 2])` in cases and `r gsub("\\\\", "", table.agegr[13, 1])` in controls.  The odds ratio for severe disease associated with encashment of at least one prescription in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).  These associations remained after excluding those with any of the listed conditions.  The associations with hospital diagnoses were distributed over most chapters of the ICD-10 and the associations with prescribed drugs were distributed over most chapters of the British National Formulary.

With a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.  The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, `r table.densities[2, 5]` bits for a model that included the listed conditions, and `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.

@@

We report a systematic examination of the health care record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the general population. The objectives were to identify risk factors and to lay the basis for risk stratification based on a predictive model.

# Methods
## Case definition 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.  The CHI number is a unique identifier used in all care systems in Scotland.  All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.  Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and from the Scottish Morbidity Record 01 which is updated weekly.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that covers all critical care units in Scotland. Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.

Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.  By restricting the case definition to cases that were fatal or received critical care, we ensured complete ascertainment of all test-positive cases that were severe enough to have been fatal without critical care, irrespective of any selection that may have excluded those assessed as unlikely to benefit from critical care.

## Matched controls
For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case using the Community Health Index (CHI) database. As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.

## Demographic data
Socioeconomic status was scored in the CHI database as the Scottish Index of Multiple Deprivation, an indicator based on postal code.  Ethnicity was assigned based on applying a name classification algorithm (ONOMAP) [@lakha_name_2011] to the names in the CHI database.  This classification is able to distinguish South Asians, the largest non-White minority in Scotland, from those who identify as White with fair accuracy but misclassifies most of those who identify as Black or Black-Caribbean (a much smaller minority in Scotland) as White. <!--- For cases and controls who had a past record of hospital admission or outpatient attendance in the Scottish Morbidity Record, a separate classification of ethnicity was available, assigned by self-report or by hospital staff and based on the categories used in the Census. --->

## Morbidity and drug prescribing
For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2 and using all codes on the discharge. British National Formulary (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive.   Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.

We began by scoring a specific list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  These listed conditions were assigned based on diagnostic codes in any hospital discharge record during the last five years, or encashed prescription of a drug for which the only indications are in that group of diagnostic codes. The ICD-10 codes and drug codes used to define each variable are listed in the R script provided as Supplementary Material. In addition to these pre-specified groupings of codes we tabulated associations with all ICD-10 codes and BNF codes to chapter level.  A
separate list of conditions has been used to designate "clinically extremely vulnerable" who have been advised to shield themselves completely since early in the epidemic: this list includes solid organ transplant recipients, people receiving chemotherapy for cancer, and people with cystic fibrosis or leukaemia.  We did not separately tabulate these conditions as we expected these individuals to be underrepresented among cases if shielding was adequate.

## Statistical methods
To estimate the relation of cumulative incidence and mortality from COVID-19 to age and sex, logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years. For the case-control study, estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.  Among those cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF chapter with severe COVID-19. Where the sample of cases and controls is restricted, this will generate strata that contain no cases.  These strata will be ignored by the conditional logistic regression model as they do not contribute to the conditional likelihood.  With incidence density sampling, the odds ratios in conditional logistic regression models are equivalent to rate ratios.  These odds ratios based on the conditional likelihood will differ from the unconditional odds ratios calculable from the frequencies of exposure in cases and controls [@breslow_estimation_1978]

To construct risk prediction models, we used stepwise regression alternating between forward and backward steps to maximize the AIC, implemented in the R function `stats::step`.  The performance of the risk prediction model as a classifier was examined by 4-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination $\Lambda$ [@mckeigue_quantifying_2019] and (more conventionally) as the C-statistic. The rationale for using $\Lambda$ to quantify the predictive performance of a binary classifier, and especially to quantify increments in predictive performance, has been explained elsewhere [@mckeigue_quantifying_2019].  A key advantage of using $\Lambda$ is that the contributions of independent predictors are additive on the scale of $\Lambda$.  The class-conditional distributions of weights of evidence in cases and controls encode exactly the same information as the receiver operating characteristic curve, when these distributions are adjusted for mathematical consistency as described elsewhere [@mckeigue_quantifying_2019] to obtain the model-based distributions.  The likelihood ratio at any given sensitivity and specificity is the gradient of the model-based ROC curve.

In this study we have estimated the information for discrimination provided by age and sex (from a logistic regression model in the general population), and the information for discrimination provided by other risk factors conditional on age and sex (from conditional logistic regression models in this case-control study matched for age and sex).  We can add these to obtain the total information for discrimination that would be obtained with a classifier in the general population.  As described elsewhere, this can be used to estimate the proportion of those destined to develop severe or fatal disease who would be correctly classified as high risk under a shielding policy [@mckeigue_evaluation_2020].

## Information governance
This study was conducted under approvals from the  Privacy Advisory Committee ref 44/13  and Public Benefit Privacy Protection amendment 1617-0147 . Datasets were de-identified before analysis.

# Results
## Incidence and mortality from severe COVID-19 in the Scottish population
Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.  The relationship of mortality to age is almost exactly linear on a logit scale, and the lines for males and females are almost exactly parallel.  For severe cases as defined in this study, the sex differential is narrow up to about age 50 but widens between ages 50 and 70 years.  This implies that at younger ages the ratio of critical care admissions to total fatalities is higher in women than in men, but that at later ages the ratio of critical care admissions to total fatalities is higher in men.  In models that included  age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(10*logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease.  The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.  As shown in Table \ref{tab:agegr} almost all severe cases over aged 75 years died but most severe cases under aged 60 survived.

## Association with demographic factors
Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.  Residence in a care home was by far the strongest risk factor for severe disease.  Higher risk of severe disease was also associated with socioeconomic deprivation.  In comparison with those classified as White, South Asians and other non-White ethnic minorities were at lower risk of severe disease.  Table \ref{tab:ethnicsmr} shows a separate analysis restricted to those with ethnicity recorded in hospital records, which allows those who identify as Black to be distinguished from those who identify as White.  The numbers of cases and controls classified as Black are too small for the risk in this group to be estimated reliably.

## Associations with listed conditions
Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of the listed conditions.  The prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.  `r gsub("\\\\", "", table.agegr[13, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[13, 1])` of the controls.  In those aged 75+ years `r gsub("\\\\", "", table.agegr[13, 8])` of the cases and `r gsub("\\\\", "", table.agegr[13, 7])` of the controls had at least one listed condition.

Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} examine these associations by age group, with the 0-39 and 40-59 age bands combined.  All listed conditions were associated with severe disease in each age band.  Diabetes was associated with a 6.6 fold rate of severe COVID-19 under age 60 years.  The associations with severe COVID-19 were stronger for the broad category "other heart disease" than for ischaemic heart disease.

Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} also show multivariate rate ratios by age group. These table show that even after adjusting for care home residence and listed conditions, encashment of any prescription in the past year and any admission to hospital in the past five years were strongly and independently associated with severe disease.

## Systematic analysis of diagnoses associated with severe disease
The approximately two-fold risk of with severe COVID-19 associated with prior hospital admission independently of other conditions and factors was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, restricted to those without any of the listed conditions.  These results are shown in Table \ref{tab:conditions}.  In univariate analyses, almost all ICD chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.  The strongest residual associations were seen with neoplasms. Table \ref{tab:icdsubchapters} extracts univariate associations with ICD subchapters where the p-value is <0.001 and where there are at least 50 cases and controls with a diagnosis in this subchapter.  This table shows that many subchapter diagnoses are associated with markedly higher risk of severe COVID-19.  We note that past hospital diagnoses of infections, pneumonia and acute respiratory diseases are strongly associated with severe COVID-19.  Associations with cardiovascular disease are not limited to heart disease but are also present for stroke and other circulatory disorders that are not designated as risk conditions.

@@

To investigate this further, we partitioned the "Any prescription" variable into indicator variables for each chapter of the British National Formulary, in which drugs are grouped by broad indication, and restricted the analysis to those without one of the listed conditions.  Table \ref{tab:drugs} shows these associations.  In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.  In a multivariate analysis of all chapters, these associations were much weaker.  The BNF chapter with the strongest association with severe disease was Chapter 1 (gastrointestinal).  Other chapters associated with severe disease were Chapter 2 (cardiovascular) and Chapter 4 (central nervous system).

@@

This estimate of `r table.densities[3, 5] ` bits for the information conditional on age and sex obtained from the matched case-control study can be added to the information for discrimination `r round(cases.Lambda.agesex, 2)` bits obtained from the logistic regression on age and sex in the population using age and sex to estimate the total information for discrimination of a risk classifier that would be obtained in the population to give a total of `r table.densities[3, 5] + round(cases.Lambda.agesex, 2)` bits.

>>>>

---
title: 'Rapid Epidemiological Analysis of Comorbidities and Treatments as risk factors in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population '
author: 
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    email: paul.mckeigue@ed.ac.uk
    affiliation: HPS
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart J McGurnaghan
    affiliation: IGMM
  - name: Sharon Kennedy
    affiliation: HPS
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Thomas M Caparrotta 
    affiliation: IGMM
  - name: Alison Smith-Palmer 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    affiliation: HPS
  - name: Public Health Scotland COVID-19 Health Protection Study Group
    affiliation: HPS
address: 
  - code: Usher
    address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care

  - code: IGMM
    address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology.

  - code: HPS
    address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE

  - code: Glasgow
    address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health

  - code: Strathclyde
    address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology

  - code: GCU
    address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health

  - code: ISD
    address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health.

geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true

csl: ../plos.csl
urlcolor: red
---

@@

@@

_Methods_ -- The design was a matched case-control study. Cases were ascertained as all those with a positive nucleic acid test for SARS-CoV-2 in the national database.  Those who had been admitted to an intensive care unit or had died within 28 days of a positive test were classified as severe cases.  For each case, ten controls matched for sex, age and primary care practice were selected from the population register.  For cases and controls we extracted all diagnostic codes from the past five years of hospital discharge records, all drug codes from encashed prescriptions issued in primary care care during the past year and linked to the national diabetes register.  Rate ratios for associations with severe COVID-19 were estimated by conditional logistic regression.  The performance of predictive models for severe COVID-19 was estimated by cross-validation.

_Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]`.  Univariate rate ratios (95% CIs) for eight conditions listed by public health agencies as conferring high risk were `r table.agegr.all[7, 3]` for Type 2 diabetes, 
`r table.agegr.all[8, 3]` for ischemic heart disease,
`r table.agegr.all[9, 3]` for other heart disease, 
`r table.agegr.all[10, 3]` for chronic lower respiratory tract disease, 
`r table.agegr.all[11, 3]` for chronic kidney disease, 
`r table.agegr.all[12, 3]` for neurological disease, 
`r table.agegr.all[13, 3]` for chronic liver disease and 
`r table.agegr.all[14, 3]` for immune deficiency or suppression.

`r freqs.all[16, 2]` of the severe cases and `r freqs.all[16, 1]` of the controls had at least one of these listed conditions.  In those aged under 40 years the frequencies of any listed condition were `r gsub("\\\\", "", table.agegr[16, 2])` in cases and `r gsub("\\\\", "", table.agegr[16, 1])` in controls.  The odds ratio for severe disease associated with encashment of at least one prescription in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).  These associations remained after excluding those with any of the listed conditions.

With a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.  The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, increasing to `r table.densities[2, 5]` bits for a model that included the listed conditions, and to `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.

@@

We report a systematic examination of the hospitalisation record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the general population. The objectives were to identify risk factors and to lay the basis for risk stratification based on a predictive model.

# Methods
## Case definition 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally.  Individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` in ECOSS were ascertained.  Using the Community Health Index (CHI) identifier contained in ECOSS (the CHI number is a unique identifier used in all care systems in Scotland) linkage to other datasets was carried out. Hospital admissions from the time of testing were obtained from the RAPID database a daily return of current hospitalisations each day.  Admissions to intensive care units (ICUs) care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that covers all admissions to all adult ICUs and to combined ICU/high dependency units ( hereafter abbreviated as ICU) across Scotland and has returned a daily census of patients in ICU from the beginning of the COVID-19 epidemic. Death registrations up to `r gsub("^0", "", format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y"))` were obtained from linkage to the National Register of Scotland.

Severe or fatal COVID-19 was defined by a record of entering ICU in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.  By restricting the case definition to cases that were fatal or received ICU care, we ensured complete ascertainment of all test-positive cases that were severe enough to have been fatal without ICU care, irrespective of any selection that may have excluded those assessed as unlikely to benefit from ICU care.

## Matched controls
For each test-positive case, we ascertained ten matched controls of the same sex, one-year age band and registered with the same primary care practice who were alive on the date of the first test in the case using the Community Health Index (CHI) database. As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.

## Demographic data
Residence in a care home was ascertained from the CHI database. Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation, an indicator based on postal code.  Ethnicity was assigned based on applying a name classification algorithm (ONOMAP) [@lakha_name_2011] to the names in the CHI database.  Cross-tabulation of a subset of records for which ethnicity assigned by patients or health care staff had been recorded in the Scottish Morbidity Record showed that this algorithm correctly classified 90% of those who reported South Asian ethnicity (1.6% of the  Scottish population in 2011) but misclassified most of those who identified as Caribbean or Black (0.1% of the population in 2011) as White.  <!--- For cases and controls who had a past record of hospital admission or outpatient attendance in the Scottish Morbidity Record, a separate classification of ethnicity was available, assigned by self-report and based on the categories used in the Census. --->

## Morbidity and drug prescribing
For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2 and using all codes on the discharge. Diagnostic coding under ICD chapters 5 (Mental, Behavioural and Neurodevelopmental) and 15 (Pregnancy) is incomplete as most psychiatric hospital and pregnancy units returns are not captured in SMR01. British National Formulary (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.  The BNF groups drugs by 2-digit chapter codes.  For this analysis prescription codes from chapters 14 and above, mostly for dressings and appliances but also including vaccines were grouped as "Other".

We began by scoring a specific list of eight conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  A separate list of conditions designates "clinically extremely vulnerable" who have been advised to shield themselves completely since early in the epidemic: this list includes solid organ transplant recipients, people receiving chemotherapy for cancer, and people with cystic fibrosis or leukaemia.  We did not separately tabulate these conditions as we expected these individuals to be underrepresented among cases if shielding was adequate.

The eight listed conditions were scored based on diagnostic codes in any hospital discharge record during the last five years, or encashed prescription of a drug for which the only indications are in that group of diagnostic codes. The R script included as supplementary material contains the derivations of these variables from ICD-10 codes and BNF drug codes.  Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other/Unknown. Cases of diabetes in the last update of the register were supplemented with new cases identified through discharge codes and drug codes.  These new cases were coded as type Other/Unknown.

## Statistical methods
To estimate the relation of cumulative incidence and mortality from COVID-19 to age and sex, logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years. To assess whether allowing for a non-linear relationship of the logit of risk to age would improve the fit, we also fitted generalized additive models, implemented in the R function `gam::gam`, with default smoothing function.

For the case-control study,  all estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.  Among those cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF chapter with severe COVID-19.  Where an exclusion criterion such as having a pre-defined condition was applied to cases  this was also applied to controls as otherwise subsequent association estimates would be incorrect. Where the sample of cases and controls is restricted, this will generate strata that contain no cases but these strata will be ignored by the conditional logistic regression model as they do not contribute to the conditional likelihood.  With incidence density sampling, the odds ratios in conditional logistic regression models are equivalent to rate ratios.  These odds ratios based on the conditional likelihood will differ from the unconditional odds ratios calculable from the frequencies of exposure in cases and controls [@breslow_estimation_1978]
Although matching on primary care practice will match to some extent for associated variables such as care home residence, socioeconomic disadvantage and prescribing practice, the effects of these variables are still estimated correctly by the conditional odds ratios.

To construct risk prediction models, we used stepwise regression alternating between forward and backward steps to maximize the AIC, implemented in the R function `stats::step`.  The performance of the risk prediction model in classifying cases versus non-cases of severe COVID-19 was examined by cross-validation. We calculated the performance calculated over all test folds using the C-statistic but also using the "expected information for discrimination" $\Lambda$ expressed in bits [@mckeigue_quantifying_2019].  The use of bits (logarithms to base 2) to quantify information is standard in information theory: one bit can be defined as the quantity of information that halves the hypothesis space.  Although readers may be unfamiliar with the expected information for discrimination, it has several properties that make it more useful than the C-statistic for quantifying the performance of a risk prediction model.  [@mckeigue_quantifying_2019] .  A key advantage of using $\Lambda$ is that contributions of independent predictors can be added.  Thus in this study we can add the predictive information from a logistic model of age and sex in the general population to the predictive information provided by other risk factors from the case-control study matched for age and sex.

## Information governance
This study was conducted under approvals from the  Privacy Advisory Committee ref 44/13  and Public Benefit Privacy Protection amendment 1617-0147. Datasets were de-identified before analysis.

# Results
## Incidence and mortality from severe COVID-19 in the Scottish population
Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.  The relationship of mortality to age is almost exactly linear on a logit scale, and the lines for male and female mortality are almost parallel.  For severe cases as defined in this study, the sex differential is narrow up to about age 50 but widens between ages 50 and 70 years.  This implies that at younger ages the ratio of ICU admissions to total fatalities is higher in women than in men, but that at later ages the ratio of ICU admissions to total fatalities is higher in men.  In models that included age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(10*logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease. The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.  As shown in Table \ref{tab:agegr} almost all severe cases aged over 75 years died but most severe cases aged under 60 years survived.

## Association with demographic factors
Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.  Residence in a care home was by far the strongest risk factor for severe disease.  Higher risk of severe disease was also associated with socioeconomic deprivation.  In comparison with those classified as White, South Asians were at lower risk of severe disease.  <!---Table \ref{tab:ethnicsmr} shows a separate analysis restricted to those with ethnicity recorded in hospital records, which allows those who identify as Black to be distinguished from those who identify as White.  The numbers of cases and controls classified as Black are too small for the risk in this group to be estimated reliably.--->

## Associations with listed conditions
Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of the listed conditions.  The prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.  `r gsub("\\\\", "", table.agegr[16, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[16, 1])` of the controls.  In those aged 75+ years `r gsub("\\\\", "", table.agegr[16, 8])` of the cases and `r gsub("\\\\", "", table.agegr[16, 7])` of the controls had at least one listed condition.

Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} examine these associations by age group, with the 0-39 and 40-59 age bands combined.  All listed conditions were associated with severe disease in each age band.  In those aged under 60 years, the rate ratio was `r tables.agegr[[1]][6, 3]` for Type 1 diabetes and `r tables.agegr[[1]][7, 3]` for Type 2 diabetes.  For all age groups combined, the rate ratio was `r tables.agegr[[1]][8, 3]` for ischemic heart disease and `r tables.agegr[[1]][9, 3]` for the broad category "other heart disease".  The multivariate analyses shown in Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that in each age group dispensing of any prescription in the past year and any admission to hospital in the past five years were strongly and independently associated with severe disease even after adjusting for care home residence and listed conditions.

## Systematic analysis of diagnoses associated with severe disease
The association of severe COVID-19 with prior hospital admission was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, restricted to those without any of the listed conditions.  These results are shown in Table \ref{tab:conditions}.  In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.  The strongest residual associations were seen with diagnoses in ICD chapters 2 (neoplasms), 5 (Mental) and 10 (musculoskeletal). Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters, filtered to show only subchapters for whcih the univariate p-value is <0.001 and there are at least 50 cases and controls with a diagnosis in this subchapter.  This was restricted to those without any listed conditions. This table shows that many subchapter diagnoses are associated with markedly higher risk of severe COVID-19.  We note that past hospital diagnoses of infections, pneumonia and acute respiratory diseases are strongly associated with severe COVID-19. Associations with cardiovascular disease are not limited to heart disease but are also present for stroke and other circulatory disorders that are not designated as risk conditions.

@@

To investigate this further, we partitioned the "Any prescription" variable into indicator variables for each chapter of the British National Formulary, in which drugs are grouped by broad indication, and restricted the analysis to those without one of the listed conditions.  Table \ref{tab:drugs} shows these associations.  In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.  In a multivariate analysis of all chapters, these associations were much weaker.  The BNF chapters with the strongest association with severe disease were chapters 1 (gastrointestinal) and 2 (cardiovascular)).  Other chapters associated with severe disease were chapters 4 (central nervous system), 9 (nutrition and blood) and 14+ (other, mostly dressings and appliances).

@@

This estimate of `r table.densities[3, 5] ` bits for the information conditional on age and sex obtained from the matched case-control study can be added to the information for discrimination `r round(cases.Lambda.agesex, 2)` bits obtained from the logistic regression on age and sex in the population using age and sex to estimate the total information for discrimination of a risk classifier that would be obtained in the population as `r table.densities[3, 5] + round(cases.Lambda.agesex, 2)` bits.
